\chapter{Potential Applications}
\label{ch:Applications}

After doing a study about the different systems to monitor and analyse the postural adjustments, we proceed to explain possible applications in the real life.

The force plate system is a very accurate system to analyse disorders in the patients, however it is a limited system for its price and portability. So, its applications make restricted to diagnosis some diseases.
The same happens with Qualisys System because it is necessary fixed cameras to record data. However it is a interesting way to observe data in real time with precision and robustness .

But, without a doubt, Gait Watch System is one of the most interesting system due to portability and its amount of fields where it can be used such as telerehabilitation, daily activities and performance of some athletics. This is why, the majority of applications will be focused in this system.

All of these implementations will be briefly explained below as well as a business idea as a concrete application of this Project.

\section{Diseases}
There exists a large amount of diseases that distort the motor control of human body or present symptoms that can be identified by the analysis of human body posture and motion. Along this section,  we will briefly comment how our study has influence in that.

\subsection{Neurological and Muscular diseases}
Parkinson’s disease is the second neurodegenerative disorder more frequent after Alzheimer\cite{AppPD}. According to the Parkinson’s disease fundation, PD is a chronic and progressive movement disorder, meaning that symptoms continue and worsen over time\cite{pdf}.
Many as one million Americans and 60,000 Spaniards and 10 million people worldwide live with Parkinson’s disease. Also, there is a large number of cases that go undetected \cite{A.Olivares2013}.

Primary motor signs of Parkinson’s disease include trembor, bradykinesia, dyskinesia and disorders in the posture and gait. Bradykinesia is the term for defining slow execution of the movement, tremor is the term used to define repetitive periodic movements within a certain frequency range and dyskinesia is a movement disorder which consists of effects including diminished voluntary movements and the presence of involuntary movements .Inertial Systems could be used to detect these disorder and help to diagnose and monitor them \cite{A.Olivares2013}.

Another disease associated to the central nervous system is Multiple Sclerosis (MS). MS is an autoinmune disease characterize bay scar tissue resulting from the repair of damage to the myelin sheath that surronunds neurons. MS affects approximately 2.5 million people worlswide. Symptoms of MS are unpredictable but may inlcude fatigue, visión problems, loss of balance and coordination, or depression. Also, Individuals with MS often have por balance control that is especially apparent during dynamic task suach as gait initation. Hence, Inertial Sensors may be a useful tool to detect these symptoms \cite{Jebb2008}.


Cerebral  palsy (CP)is important to be mention  because it is a movement disorder appears in early childhood. CP is a neurodevelopmental condition caused by non-progressive brain lesión, can occur before, during or shortly after birth. Children diagnosed with CP demonstred increased muscle activity to sustain posture, agonist/antagonist co-contraction, impair postural control, inadecuate force production, and restrictive voluntary ans selective control of movement\cite{Gay2011}.

These impairments nor only interfere with performance of functional activities, but also with opportunities  ans/or willingness to participate in leisure, community ans achool activities.

Impaired postural control in CP includes difficulty organizing compensatory postural adjustments (CPAs) and anticipatory postural adjustments (APAs) \cite{Gay2011}. For that purpose,  consideration should be given to the use of Inertial Sensors to monitor this disease in dialy life.

\subsection{Sleep disorder}
Sleep disorders cause an unrestful sleep and important repercussions in some cases such as sleepiness and psychiatric and cardiorespiratory secondary disorders\cite{SanchezDaniel}. 

In order to diagnose them, we can use inertial sensors to monitor changes that occur during sleep. In addition, it is a good option to do this  cheap and portable, in such a way that patient can move freely while they are being monitored\cite{A.Olivares}.

Also, the information gathered can provide information  of the cardiac, respiratory, and snoring activities of patients sleeping\cite{SanchezDaniel}.

If we use all of this at home, it is possible to provide a tool to sleep specilists for knowing the behaviour of the patient when they are sleeping, what sleep cycles are more affected and improve their medical  treatments. 

For example,  some of the patients with Epilepsy suffer as tonic-clonic nocturnal seizures. These seizures  may go unnoticed to the patient, thus, during the control sessions with the doctor the patient could not tell the doctor these episodes. To avoid such a situation, the patient may sleep with an attached MIMU so nocturnal seizures can be detected and stored in the memory, allowing the doctor to notice that nocturnal attacks are happening\cite{A.Olivares2013}. 

\section{Dialy activities }
Providing new wereable technology for medical and surgical rehabilitation services is emerging as important option for clinicians and patients. Wereable technology like inertial sensors provides  a convenient platform to be able to quantify the long-term context and physiological response for  individuals\cite{Sung}.

In the first phase of the reabilitation program the patient have to move to the medical center for treatment. However, the second phase patient have to do low intensity exercices to stregthen  the muscles. This last task doesn’t require continuous  supervisión, thus, MARG systems can avoid the overcrowding of rehabilitation centers and patient can do the exercise at home comfortably\cite{A.Olivares2013}.

In addition, doctors can then supervise the sessions carried out by the patients by remotely checking the logs. Also, they could detect differents activity states, for example, knowing whether a person  is sleeping, driving or doing exercise\cite{A.Olivares2013}.

Fall detection is a important aspect during daily life of elderly people . It is very clear that falls are a serious health issue and that systems automatically detecting a fall and calling for help could be of great help to solve it. 

Moreover, there are applications for the motion analysis in sport activities by attaching sensors like accelerometers and gyroscopes onto the athletes ‘  body segments. For example, for a swimmer, the discrimination of the swimming styles and the segmentation of the underwater stroke phases could be achieved. In addition, the physiological response was also detected on the wrist acceleation when the swimmer was fatigued in analsis method the intensive training situation\cite{Yuji}.



\section{Business plan }