\chapter{Conclusion and Future Work}
\label{ch:Conclusion and Future Work}

\section{Conclusion}

Gait analysis is a useful tool both in clinical practice and biomechanical research. In this work an extended Kalman filter based on a kinematic model of the leg was implemented in \textsc{Matlab}\textsuperscript{\textregistered}, along with a parameter optimiser that was necessary for testing. Then, the filter algorithm was applied to movement data of a human subject walking on a treadmill at different speeds. The experiments have shown that the filter algorithm can improve the angle estimates by an average RMSE that was $28.52\%$ smaller, compared to the classical Kalman filter.
However, the motion-based acceleration correction works better for faster motion. In order to replace a camera-based motion capture system with low cost wearable MARG sensors, some technical details still need to be improved. Numerous possible reasons for the behaviour seen in the results of the experiments were proposed. These include a lack of completeness of the mathematical model of the leg, unknown lengths of the thighs and shanks, and a non-optimal filter tuning.

Personally, I have learned a lot in the seven months that I spent in Granada for writing this thesis and completing the internship in advance. Among others, I have come to know many new work methods, due to being exposed to people from different cultures. I gained experience in scientific research and a deeper understanding of human body motion analysis, in particular gait analysis, as well as how Kalman filtering benefits the accuracy of the orientation estimation. I improved my \textsc{Matlab}\textsuperscript{\textregistered} skills and I am now familiar with tools such as GitHub and Pivotal Tracker, which make working in a team much easier and significantly more efficient.  While working at the research centre, I could improve my oral and written English skills.

All in all it was a great experience, professionally as well as personally. I truly and unreservedly recommend such a stay to \emph{every single} university student.

\section{Future Work}

As discussed above, there are a variety of possible improvements regarding the kinematic model and the filter tuning, which lead to the following possible future work. One could extend the kinematic model of the leg, in order to take more complex movements of the body and especially the movement of the hip into account. Furthermore, the implemented orientation algorithm needs to be tested with a larger set of data from real subjects, in order to proof its validity and robustness. Especially, whether it delivers accurate orientation estimations without fine tuning the parameters by means of an optimiser and a known reference for every patient and trial, needs to be tested. Finally, the acceleration correction based on a kinematic model could be applied to the accelerometer-based thigh angle estimate as well.

Regarding the GaitWatch system itself, the acceleration correction could be extended to the orientation estimation of the arms and the trunk. Then, the GaitWatch system could replace the camera-based motion capture systems in the future and would thus add a significant value to its medical applications in gait analysis.


